---
title: "Process data"
author: "Stefano Coretta"
format: html
---

```{r}
#| label: setup

library(tidyverse)
library(phonetisr)
library(speakr)
```


## Forced alignment

The text files exported from AAA contain the sentence on the first line, so we remove all the other lines from the text files.

```{bash}
folder="raw/audio"

cd "$folder" || exit

for file in *.txt; do
  first_line=$(head -n 1 "$file")

  echo "$first_line" > "$file"
done
```

Now we can validate the corpus.

```bash
mfa validate --clean --single_speaker data-raw/raw/audio basque_cv
```

Let's transcribe the OOVs.

```{r}
oovs <- read_lines("raw/oovs_found_basque_cv.txt")

repl <- c(
  "cq" = "k",
  "dd" = "ɟ",
  "tt" = "c",
  "rr" = "R",
  "ll" = "ʎ",
  "d" = "d̪",
  "t" = "t̪",
  "s" = "s̺",
  "g" = "ɡ",
  "ñ" = "ɲ",
  "r" = "ɾ",
  "R" = "r",
  "x" = "ʃ",
  "j" = "x",
  "y" = "j",
  "ü" = "y",
  "z" = "s̻",
  "q" = "k",
  "ç" = "s",
  "h" = ""
)

oovs_tra <- tibble(
  word = str_to_lower(oovs)
) |> 
  arrange(word) |> 
  distinct() |> 
  mutate(
    pron = str_replace_all(word, repl),
    ipa = phonetise(pron, split = FALSE, sep = " ", diacritics = TRUE)
  )

write_tsv(oovs_tra |> select(word, ipa), "raw/oovs_transcribed.txt", col_names = FALSE)

```

And we can now add them to the MFA dictionary.

```bash
mfa model add_words basque_cv data-raw/raw/oovs_transcribed.txt
```

We can finally force-align the audio files.

```bash
mfa align --clean --verbose --single_speaker --fine_tune data-raw/raw/audio basque_cv basque_cv data-raw/raw/aligned/
```

## Get UTI data

First, we process the TextGrids to extract only the phones from the target words, which we will use in AAA to extract spline data.

```{r}
praat_run("praat/01_extract_target.praat")
```

The new TextGrids are saved in `data-raw/raw/aligned_target/`. To import them in AAA, they need to be in the same folder as the `.txt` files exported from AAA (the original ones, with all the lines, not just the first).

We extract all ultrasound frames from vowels to `data-raw/raw/splines/vowels.tsv`.
